---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---
# Final Assignment - Practical Machine Learning - Coursera
## by Simon J. Baumgart 
## Introduction and background 
### Objective:
One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal was to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants and _predict_ how well they were doing the excercize compared to different compared movements:  

* exactly according to the specification (Class A), 
* throwing the elbows to the front (Class B)
* lifting the dumbbell only halfway (Class C) 
* lowering the dumbbell only halfway (Class D) 
* throwing the hips to the front (Class E)

## Review Criteria:
The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

### Background:
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

### Data & Material
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

Standard laptop PC for model prediction.

## Results
### data exploration & identification of important predictor variables

```{r include = F, echo=F, results='hide', eval=T}
library(caret)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(mlbench)
library(parallel)
library(doParallel)
library(ggfortify)
library(rfUtilities)
```

```{r include = F, echo=F, results='hide', eval=T}
#download.file(url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "~/Data_Science/course 8 - practical machine learning/data/pml-training.csv")
#download.file(url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "~/Data_Science/course 8 - practical machine learning/data/pml-testing.csv")
trainingdata <- read_csv("~/Data_Science/course 8 - practical machine learning/data/pml-training.csv")
trainingdata.master <- trainingdata
inTrain <- createDataPartition(y = trainingdata.master$classe, p = 0.85, list = F)

trainingdata <- trainingdata.master[inTrain,]
crossvaliddata <- trainingdata.master[-inTrain,] 

testingdata <- read_csv("~/Data_Science/course 8 - practical machine learning/data/pml-testing.csv")
```

```{r include = F, echo=F, results='hide', eval=T}
trainingdata <- read_csv("~/Data_Science/course 8 - practical machine learning/data/pml-training.csv")
testingdata <- read_csv("~/Data_Science/course 8 - practical machine learning/data/pml-testing.csv")
```

### dimension of the datasets
```{r include = F, echo=F, results='show', eval=T}
dim(testingdata)
dim(trainingdata)
dim(crossvaliddata)
```

```{r include = F, echo=F, results='hide', eval=T}
na_list <- c()
for(i in  1:ncol(trainingdata)){if(sum(is.na(trainingdata[,i])) > nrow(trainingdata)*2/3){
    na_list <<- append(na_list,i)
    
}
}

trainingdata_NArm <- trainingdata[,-(na_list)]

# I remove timestamp part 2 and cvtd timestamp as the inforamtion is already in timestamp_part_1
trainingdata_NArm_tidy <- trainingdata_NArm[,-c(1,4,5)]
```


```{r identification  removal of an outlier, include = F, echo=F, results='hide', eval=T }
#install.packages("ggfortify")
trainingdata_NArm_tidy_PCA <- trainingdata_NArm_tidy
trainingdata_NArm_tidy_PCA$user_name <- as.numeric(as.factor(trainingdata_NArm_tidy_PCA$user_name))
trainingdata_NArm_tidy_PCA$new_window <- as.numeric(as.factor(trainingdata_NArm_tidy_PCA$new_window))

trainingdata_NArm_tidy_PCA$classe <- as.numeric(trainingdata_NArm_tidy_PCA$classe)

glimpse(trainingdata_NArm_tidy_PCA)

glimpse(trainingdata_NArm_tidy_PCA[,-c(ncol(trainingdata_NArm_tidy_PCA))])
trainingdata_NArm_tidy.pca <- prcomp(trainingdata_NArm_tidy_PCA[,-c(ncol(trainingdata_NArm_tidy_PCA))], center = T, scale. = T)

autoplot(trainingdata_NArm_tidy.pca,data = trainingdata_NArm_tidy, colour = 'user_name', label = TRUE, label.size = 5)
#5373 as outlier in eurico to remove

autoplot(trainingdata_NArm_tidy.pca,data = trainingdata_NArm_tidy, colour = 'user_name', loadings = TRUE, loadings.colour = 'blue',
loadings.label = TRUE, loadings.label.size = 3, x=3, y =4)

autoplot(cluster::clara(trainingdata_NArm_tidy_PCA[,-c(ncol(trainingdata_NArm_tidy_PCA))],6))
## this indicates that the users have a huge implication on the variance 

#remove outlier
trainingdata_NArm_tidy.outRM <- trainingdata_NArm_tidy[-5373,]
trainingdata_NArm_tidy_PCA.outRM <- trainingdata_NArm_tidy_PCA
trainingdata_NArm_tidy_PCA.outRM <- trainingdata_NArm_tidy_PCA.outRM[-5373,]

trainingdata_NArm_tidy_PCA.outRM.pca <- prcomp(trainingdata_NArm_tidy_PCA.outRM[,-c(ncol(trainingdata_NArm_tidy_PCA.outRM))], center = T, scale. = T)
```

#### pca plot 
__plot1__  
The data is mainly distributed by the training individuals looking at the different PC    

```{r include = T, echo=F , fig.height = 2, fig.width = 5}
autoplot(trainingdata_NArm_tidy_PCA.outRM.pca ,data = trainingdata_NArm_tidy.outRM, colour = 'user_name')
autoplot(trainingdata_NArm_tidy_PCA.outRM.pca ,data = trainingdata_NArm_tidy.outRM, colour = 'user_name', x = 2, y = 3)
autoplot(trainingdata_NArm_tidy_PCA.outRM.pca ,data = trainingdata_NArm_tidy.outRM, colour = 'user_name', label.size = 5, x = 3, y =4)
```

__plot2__  
When plotting the different "classe" variables it appears that there is __no__ clear difference between the "classes" and that they get distributed well. They rather get distributed by individuals who performed the test. Therefore, I would like to identify key variables. 

```{r include = T, echo=F , fig.height = 2, fig.width = 5}
autoplot(trainingdata_NArm_tidy_PCA.outRM.pca ,data = trainingdata_NArm_tidy.outRM, colour = 'classe')
autoplot(trainingdata_NArm_tidy_PCA.outRM.pca ,data = trainingdata_NArm_tidy.outRM, colour = 'classe', x = 2, y = 3)
autoplot(trainingdata_NArm_tidy_PCA.outRM.pca ,data = trainingdata_NArm_tidy.outRM, colour = 'classe', label.size = 5, x = 3, y =4)
```

### reducing the number variables
Because there are many variables (160) in the dataset and the model prediction runs on a normal PC the variable reduction will save time and might reduce the risk of overfitting. 

__plot3__ variable selection out of the dataset  
  
  
```{r include = T, echo=F , fig.height = 2, fig.width = 5}
res.pca <- PCA(trainingdata_NArm_tidy_PCA.outRM[,-c(ncol(trainingdata_NArm_tidy_PCA.outRM))], graph = FALSE)
var <- get_pca_var(res.pca)
fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 50))

fviz_cos2(res.pca, choice = "var", axes = 1:5)
```
  
The cos2 indicates how much a variable is contributing to explain the variance in dimensions within the PC. I selected here for the __top 30 variables__ explaining the most variability within the first 5 dimensions of the PCA.  


```{r include = F, echo=F, results='hide', eval=T}
contributors <- (var$contrib)
contributors.df <- as.data.frame(apply(contributors, 1, sum))
contributors.df$names <-  rownames(contributors.df ) 
contributors.df.ordered <- contributors.df[order(contributors.df$`apply(contributors, 1, sum)`, decreasing = T),]
```

```{r include = F, echo=F, results='hide', eval=T}
patternmatch <- c(contributors.df.ordered[1:30,"names"],"classe")

trainingdata_NArm_tidy.outRM.selected  <- trainingdata_NArm_tidy.outRM[,grepl(colnames(trainingdata_NArm_tidy.outRM), pattern =paste(patternmatch, collapse = "|"))]
head(trainingdata_NArm_tidy.outRM)
head(trainingdata_NArm_tidy.outRM.selected)
dim(trainingdata_NArm_tidy.outRM.selected)
```

### fitting the model to the test data by a random forest
```{r modFit}
x <- as.data.frame(trainingdata_NArm_tidy.outRM.selected[,-c(1,which(colnames(trainingdata_NArm_tidy.outRM.selected) == "classe"))])
y <- trainingdata_NArm_tidy.outRM.selected$classe

fitControl <- trainControl(method = "cv",
                           number = 5,
                           allowParallel = TRUE)

cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)

mod2 <- train(x,y,method="rf", data=trainingdata_NArm_tidy.outRM.selected, trControl = fitControl, ntree=25)

stopCluster(cluster)
registerDoSEQ()
```
__Explanation why I choose this model:__  
I use parallel computing to accelerate the process. The Random Forest method is used because it is a powerful method as in the lectures explained and might be advantageous over a "glm" or "lm". I use here number of trees: 25. In addition random forest has been also used in the publication underlying this data. To improve our prediction I am using here a 5 fold cross-validation resampling technique. As an alternative bootstrapping technique could have been used. 

[why to use parallel computing in this assignment ](https://github.com/lgreski/datasciencectacontent/blob/master/markdown/pml-randomForestPerformance.md)

```{r}
mod2$finalModel
```
The out-of-bag error (OOB), which is the estimate of the error rate is with under 1% reasonable. The most difficult class to predict is the class E which is the "throwing the hips to the front". We however get a good and low class error for the "correct" exercise which is class A.


__crossvalidation on a training subset split at the beginning__ (which is not included in the model building of course)
```{r include = T, echo=T, eval=T}
crossvalidpredict.select <- crossvaliddata[,grepl(colnames(crossvaliddata), pattern =paste(patternmatch, collapse = "|"))]
crossvalidpredict <- predict(mod2, crossvalidpredict.select)
table(crossvalidpredict.select$classe,crossvalidpredict)
accuracy(crossvalidpredict.select$classe,crossvalidpredict)[1]
```
  
__The evaluation of the test set__
```{r include = T, echo=T, eval=T}
testing.selected  <- testingdata[,grepl(colnames(testingdata), pattern =paste(patternmatch, collapse = "|"))]
pred <- predict(mod2, testing.selected)
pred
```

## Conclusion 
The random forest model used here predicts  well on a split sample set at the beginning and as well on the training set. It will be thrilling to see what the outcome is on the test dataset. The hardest to predicted class in our model is class E. The best prediction with the lowest class error rate we get with the class A. This is desirable because we want to know how to train best! 

Thank you for your time and grading. 

## Apendix

### What to submit:
Your submission for the Peer Review portion should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).


## code book


```{r}
#download.file(url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "~/Data_Science/course 8 - practical machine learning/data/pml-training.csv")
#download.file(url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "~/Data_Science/course 8 - practical machine learning/data/pml-testing.csv")
trainingdata <- read_csv("~/Data_Science/course 8 - practical machine learning/data/pml-training.csv")
trainingdata.master <- trainingdata
inTrain <- createDataPartition(y = trainingdata.master$classe, p = 0.85, list = F)

trainingdata <- trainingdata.master[inTrain,]
crossvaliddata <- trainingdata.master[-inTrain,] 

testingdata <- read_csv("~/Data_Science/course 8 - practical machine learning/data/pml-testing.csv")
```

```{r include = F, echo=F, results='hide', eval=T}
library(caret)
library(ggplot2)
library(tidyr)
library(tidyverse)
library("FactoMineR")
library("factoextra")
library(mlbench)
library(parallel)
library(doParallel)
library(ggfortify)
library(rfUtilities)
```

Too many NA in the data, first step is to check which variables have lots of NA

```{r include = T, echo=T, results='hide', eval=F}
na_list <- c()
for(i in  1:ncol(trainingdata)){if(sum(is.na(trainingdata[,i])) > nrow(trainingdata)*2/3){
    na_list <<- append(na_list,i)
    
}
}

trainingdata_NArm <- trainingdata[,-(na_list)]

# I remove timestamp part 2 and cvtd timestamp as the inforamtion is already in timestamp_part_1
trainingdata_NArm_tidy <- trainingdata_NArm[,-c(1,4,5)]
```


```{r include = T, echo=T, results='hide', eval=F}
#install.packages("ggfortify")
trainingdata_NArm_tidy_PCA <- trainingdata_NArm_tidy
trainingdata_NArm_tidy_PCA$user_name <- as.numeric(as.factor(trainingdata_NArm_tidy_PCA$user_name))
trainingdata_NArm_tidy_PCA$new_window <- as.numeric(as.factor(trainingdata_NArm_tidy_PCA$new_window))

trainingdata_NArm_tidy_PCA$classe <- as.numeric(trainingdata_NArm_tidy_PCA$classe)

glimpse(trainingdata_NArm_tidy_PCA)

glimpse(trainingdata_NArm_tidy_PCA[,-c(ncol(trainingdata_NArm_tidy_PCA))])
trainingdata_NArm_tidy.pca <- prcomp(trainingdata_NArm_tidy_PCA[,-c(ncol(trainingdata_NArm_tidy_PCA))], center = T, scale. = T)

autoplot(trainingdata_NArm_tidy.pca,data = trainingdata_NArm_tidy, colour = 'user_name', label = TRUE, label.size = 5)
#5373 as outlier in eurico to remove

autoplot(trainingdata_NArm_tidy.pca,data = trainingdata_NArm_tidy, colour = 'user_name', loadings = TRUE, loadings.colour = 'blue',
loadings.label = TRUE, loadings.label.size = 3, x=3, y =4)

autoplot(cluster::clara(trainingdata_NArm_tidy_PCA[,-c(ncol(trainingdata_NArm_tidy_PCA))],6))
## this indicates that the users have a huge implication on the variance 

#remove outlier
trainingdata_NArm_tidy.outRM <- trainingdata_NArm_tidy[-5373,]
trainingdata_NArm_tidy_PCA.outRM <- trainingdata_NArm_tidy_PCA
trainingdata_NArm_tidy_PCA.outRM <- trainingdata_NArm_tidy_PCA.outRM[-5373,]

trainingdata_NArm_tidy_PCA.outRM.pca <- prcomp(trainingdata_NArm_tidy_PCA.outRM[,-c(ncol(trainingdata_NArm_tidy_PCA.outRM))], center = T, scale. = T)
```

pca plot 
the aim is to see first patterns within the data and how the variance is distributed:
```{r include = T, echo=T, results='hide', eval=F}
autoplot(trainingdata_NArm_tidy_PCA.outRM.pca ,data = trainingdata_NArm_tidy.outRM, colour = 'user_name')
autoplot(trainingdata_NArm_tidy_PCA.outRM.pca ,data = trainingdata_NArm_tidy.outRM, colour = 'user_name', x = 2, y = 3)
autoplot(trainingdata_NArm_tidy_PCA.outRM.pca ,data = trainingdata_NArm_tidy.outRM, colour = 'user_name', label.size = 5, x = 3, y =4)
```
Interpretation: 
The user has a huge influence on the variance 

```{r include = T, echo=T, results='hide', eval=F}
autoplot(trainingdata_NArm_tidy_PCA.outRM.pca ,data = trainingdata_NArm_tidy.outRM, colour = 'classe')
autoplot(trainingdata_NArm_tidy_PCA.outRM.pca ,data = trainingdata_NArm_tidy.outRM, colour = 'classe', x = 2, y = 3)
autoplot(trainingdata_NArm_tidy_PCA.outRM.pca ,data = trainingdata_NArm_tidy.outRM, colour = 'classe', label.size = 5, x = 3, y =4)
```

```{r include = T, echo=T, results='hide', eval=F}
contributors <- (var$contrib)
contributors.df <- as.data.frame(apply(contributors, 1, sum))
contributors.df$names <-  rownames(contributors.df ) 
contributors.df.ordered <- contributors.df[order(contributors.df$`apply(contributors, 1, sum)`, decreasing = T),]
```

```{r include = T, echo=T, results='hide', eval=F}
patternmatch <- c(contributors.df.ordered[1:10,"names"],"classe")
trainingdata_NArm_tidy.outRM.selected  <- trainingdata_NArm_tidy.outRM[,grepl(colnames(trainingdata_NArm_tidy.outRM), pattern =paste(patternmatch, collapse = "|"))]
trainingdata_NArm_tidy.outRM.selected$classe <- as.numeric(as.factor(trainingdata_NArm_tidy.outRM.selected$classe))
```


fitting the model - "random forest"
```{r include = T, echo=T, results='hide', eval=F}
#trainingdata_NArm_tidy.outRM.selected <- sample_n(trainingdata_NArm_tidy.outRM.selected, 10000)
# set up training run for x / y syntax because model format performs poorly
x <- as.data.frame(trainingdata_NArm_tidy.outRM.selected[,-c(1,which(colnames(trainingdata_NArm_tidy.outRM.selected) == "classe"))])
y <- trainingdata_NArm_tidy.outRM.selected$classe


fitControl <- trainControl(method = "cv",
                           number = 5,
                           allowParallel = TRUE)


cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)

mod2 <- train(x,y,method="rf", data=trainingdata_NArm_tidy.outRM.selected, trControl = fitControl, ntree=10)


stopCluster(cluster)
registerDoSEQ()
```


```{r include = T, echo=T, results='hide', eval=F}
crossvalidpredict.select <- crossvaliddata[,grepl(colnames(crossvaliddata), pattern =paste(patternmatch, collapse = "|"))]
crossvalidpredict <- predict(mod2, crossvalidpredict.select)
table(crossvalidpredict.select$classe,crossvalidpredict)
accuracy(crossvalidpredict.select$classe,crossvalidpredict)[1]
```

```{r include = T, echo=T, results='hide', eval=F}
testing.selected  <- testingdata[,grepl(colnames(testingdata), pattern =paste(patternmatch, collapse = "|"))]
pred <- predict(mod2, testing.selected)
table(testing.selected$,pred)
```

### helpful links and explanations
http://www.sthda.com/english/wiki/print.php?id=225
http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/#eigenvalues-variances
https://cran.r-project.org/web/packages/ggfortify/vignettes/plot_pca.html
https://github.com/lgreski/datasciencectacontent/blob/master/markdown/pml-randomForestPerformance.md
